---
output: github_document
---

```{r echo=FALSE, results = 'asis'}
pkg <- "rBLAST"

source("https://raw.githubusercontent.com/mhahsler/pkg_helpers/main/pkg_helpers.R")
pkg_title(pkg)
```

Interfaces the Basic Local Alignment Search Tool (BLAST) to search genetic sequence data bases with the Bioconductor infrastructure. This includes
interfaces to `blastn`, `blastp`, `blastx`, and `makeblastdb`.
The BLAST software needs to be downloaded and installed separately.

Other R interfaces for bioinformatics are also available:

* [rRDP](https://mhahsler.r-universe.dev/ui#package:rRDP): Interface to the RDP Classifier 
* [rMSA](https://mhahsler.r-universe.dev/ui#package:rMSA): Interface for Popular Multiple 
    Sequence Alignment Tools including ClustalW, MAFFT, MUSCLE, and Kalign


## Installation

1. Install the Bioconductor package `Biostrings` following the instructions [here](https://bioconductor.org/packages/release/bioc/html/Biostrings.html).
2. Install `rBlast` from r-universe using 
   ```{r, eval = FALSE}
install.packages("rBLAST", repos = "https://mhahsler.r-universe.dev")
   ```
3. Install the BLAST software by following the instructions in the 
[INSTALL](https://github.com/mhahsler/rBLAST/blob/devel/INSTALL) file.

## Usage

```{r, message=FALSE}
library(rBLAST)
```

Download the 16S Microbial data base from NCBI.

```{r, eval = FALSE}
download.file("https://ftp.ncbi.nlm.nih.gov/blast/db/16S_ribosomal_RNA.tar.gz",
    "16S_ribosomal_RNA.tar.gz",
    mode = "wb"
)

untar("16S_ribosomal_RNA.tar.gz", exdir = "16S_rRNA_DB")
```

Load the downloaded BLAST database.

```{r}
bl <- blast(db = "./16S_rRNA_DB/16S_ribosomal_RNA")
bl
```

Load some test sequences shipped with the package.

```{r}
seq <- readRNAStringSet(system.file("examples/RNA_example.fasta",
    package = "rBLAST"
))
seq
```

Query the BLAST database to find matches for the first test sequence
with a 99% percent identity or higher.

```{r}
cl <- predict(bl, seq[1, ], BLAST_args = "-perc_identity 99")
cl
```

## Citation Request

Cite the use of this package as:

> Hahsler M, Nagar A (2019). rBLAST: R Interface for the Basic Local Alignment Search Tool. 
> R package version 0.99.2, URL: https://github.com/mhahsler/rBLAST.

BibTeX
```  
@Manual{,
    title = {{rBLAST: R Interface for the Basic Local Alignment Search Tool}},
    author = {Michael Hahsler and Anurag Nagar},
    year = {2019},
    note = {R package version 0.99.2},
    url = {https://github.com/mhahsler/rBLAST}
  }
```
