---
output: github_document
---

# rBLAST - R Interface for the Basic Local Alignment Search Tool (BLAST)

```{r echo=FALSE, results = 'asis'}
pkg <- 'arules'

source("https://raw.githubusercontent.com/mhahsler/pkg_helpers/main/pkg_helpers.R")
pkg_title(pkg)
```


[![rBLAST r-universe
status](https://mhahsler.r-universe.dev/badges/rBLAST)](https://mhahsler.r-universe.dev/ui#package:rBLAST)

Interfaces the Basic Local Alignment Search Tool (BLAST) to search genetic sequence data bases with the Bioconductor infrastructure. This includes
interfaces to `blastn`, `blastp`, `blastx`, and `makeblastdb`.
The BLAST software needs to be downloaded and installed separately.

Other R interfaces for bioinformatics are also available:

* [rRDP](https://mhahsler.r-universe.dev/ui#package:rRDP): Interface to the RDP Classifier 
* [rMSA](https://mhahsler.r-universe.dev/ui#package:rMSA): Interface for Popular Multiple 
    Sequence Alignment Tools including ClustalW, MAFFT, MUSCLE, and Kalign


## Installation

1. Install the Bioconductor package `Biostrings` following the instructions [here](https://bioconductor.org/packages/release/bioc/html/Biostrings.html).
2. Install the `rBlast` from r-universe using 
   ```{r, eval = FALSE}
   install.packages('rBLAST', repos = 'https://mhahsler.r-universe.dev')
   ```
3. Install the BLAST software by following the instructions found in 
   ```{r, eval = FALSE}
   library('rBLAST')
   ? blast
   ```

## Usage

```{r, message=FALSE}
library(rBLAST)
```

Download the 16S Microbial data base from NCBI

```{r, eval = FALSE}
download.file("https://ftp.ncbi.nlm.nih.gov/blast/db/16S_ribosomal_RNA.tar.gz",
   "16S_ribosomal_RNA.tar.gz", mode='wb')

untar("16S_ribosomal_RNA.tar.gz", exdir="16S_rRNA_DB")
```

Load some test sequences

```{r}
seq <- readRNAStringSet(system.file("examples/RNA_example.fasta",
                        package="rBLAST"))
seq
```

Load a BLAST database (replace db with the location + name of the BLAST DB)

```{r}
bl <- blast(db="./16S_rRNA_DB/16S_ribosomal_RNA")
bl
```

Query a sequence using BLAST and find sequences with a 99% percent identity or higher.

```{r}
cl <- predict(bl, seq[1,], BLAST_args = "-perc_identity 99")
cl
```

## Citation Request

Cite the use of this package as:

> Hahsler M, Nagar A (2019). rBLAST: R Interface for the Basic Local Alignment Search Tool. 
> R package version 0.99.2, URL: https://github.com/mhahsler/rBLAST.

BibTeX
```  
@Manual{,
    title = {{rBLAST: R Interface for the Basic Local Alignment Search Tool}},
    author = {Michael Hahsler and Anurag Nagar},
    year = {2019},
    note = {R package version 0.99.2},
    url = {https://github.com/mhahsler/rBLAST}
  }
```
