---
title: "rBLAST: R Interface for the Basic Local Alignment Search Tool"
author: "Michael Hahsler and Anurag Nagar"
vignette: >
  %\VignetteIndexEntry{Searching Sequence Databases with rBLAST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette:
    toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup, message = FALSE}
library("rBLAST")
library("Biostrings")
```

# Overview

rBLAST connects Bioconductor sequence containers to the NCBI BLAST+
command-line programs. A typical local search has three steps:

1. Create a database from FASTA sequences with `makeblastdb()`.
2. Open the database with `blast()`.
3. Search it with `predict()` using a `DNAStringSet`, `RNAStringSet`, or
   `AAStringSet` query.

`predict()` returns a data frame with one row per BLAST hit. rBLAST can also
open downloaded NCBI databases and submit remote searches to NCBI.

# System requirements

BLAST+ is external software and must be installed separately. See the package
[installation instructions](https://github.com/mhahsler/rBLAST/blob/devel/INSTALL)
and the
[BLAST+ installation guide](https://www.ncbi.nlm.nih.gov/books/NBK569861/).

For the nucleotide workflow below, R must find all three programs:

```{r check-tools}
required_tools <- c("blastn", "makeblastdb", "blastdbcmd")
Sys.which(required_tools)
```

```{r include=FALSE}
run_local <- has_blast()
```

```{r missing-blast-note, echo = FALSE, eval = !run_local, results = "asis"}
cat(paste0(
    "**Note:** A complete local BLAST+ installation was not available when ",
    "this vignette was built. The code is shown, but output from BLAST-dependent ",
    "chunks is omitted.\n"
))
```

An empty path means that the corresponding executable is unavailable. If
BLAST+ is installed outside the normal search path, update `PATH` before using
rBLAST:

```{r set-path, eval = FALSE}
Sys.setenv(
    PATH = paste(
        Sys.getenv("PATH"),
        "/path/to/ncbi-blast/bin",
        sep = .Platform$path.sep
    )
)
```

Use `blast_help("blastn")` or run `system2("blastn", "-version")` to inspect
the installed command-line program.

# Quick start: build and query a local database

The following example is self-contained. It uses five RNA sequences bundled
with rBLAST and writes all generated files to a unique temporary directory.

```{r load-sequences}
queries <- readRNAStringSet(system.file(
    "examples/RNA_example.fasta",
    package = "rBLAST"
))
queries
```

First, write the sequences to FASTA and create a nucleotide database. RNA and
DNA sequences both use `dbtype = "nucl"`.

```{r create-database, eval = run_local}
work_dir <- tempfile("rBLAST-vignette-")
dir.create(work_dir)
fasta <- file.path(work_dir, "sequences.fasta")
db_path <- file.path(work_dir, "database")

writeXStringSet(queries, fasta)
makeblastdb(
    fasta,
    db_name = db_path,
    dbtype = "nucl",
    verbose = FALSE
)
```

`db_path` is the common prefix of the database files, not the name of one
individual index file. Open that prefix with `blast()`:

```{r open-database, eval = run_local}
db <- blast(db_path, type = "blastn")
print(db, info = FALSE)
```

Now search for a 100-nucleotide fragment taken from the first database
sequence:

```{r search-fragment, eval = run_local}
fragment <- subseq(queries[1], start = 101, end = 200)
hits <- predict(db, fragment)
hits
```

The most commonly used result columns are:

| Column | Meaning |
|:-------|:--------|
| `qseqid`, `sseqid` | Query and database-sequence identifiers |
| `pident` | Percentage of identical positions in the alignment |
| `length` | Alignment length |
| `qstart`, `qend` | Alignment coordinates in the query |
| `sstart`, `send` | Alignment coordinates in the database sequence |
| `evalue` | Expected number of chance matches; smaller is stronger |
| `bitscore` | Normalized alignment score; larger is stronger |

In this example, `sseqid` identifies the sequence from which the fragment was
taken, and `sstart` and `send` locate the fragment within that sequence.

# Multiple queries and custom output

An `XStringSet` can contain one or many query sequences. BLAST options are
passed in command-line form through `BLAST_args`. Output fields are selected
with space-separated BLAST field specifiers in `custom_format`:

```{r multiple-queries, eval = run_local}
selected_hits <- predict(
    db,
    queries[1:2],
    BLAST_args = "-perc_identity 99",
    custom_format = paste(
        "qseqid sseqid pident length",
        "qstart qend sstart send evalue bitscore"
    )
)
selected_hits
```

A successful search without matches returns a zero-row data frame. A failed
BLAST command instead produces an R error containing the executable's exit
status and the query and output paths.

# Threads and parallel searches

Use BLAST's `-num_threads` option to parallelize one local search:

```{r threads, eval = FALSE}
hits <- predict(
    db,
    queries,
    BLAST_args = "-num_threads 8"
)
```

This differs from running several `predict()` calls concurrently. Each
concurrent call creates its own query and output files and starts a separate
BLAST process. Account for both levels of concurrency when choosing worker and
thread counts; for example, 8 workers with 8 BLAST threads can request up to 64
CPU threads.

# Temporary files and troubleshooting

By default, rBLAST removes a prediction's temporary FASTA and output files.
Retain them for one diagnostic run with:

```{r debugging, eval = run_local}
hits <- predict(
    db,
    queries[1],
    verbose = TRUE,
    keep_tmp = TRUE
)

list.files(tempdir())
```

`verbose = TRUE` prints the generated filenames and BLAST command.
`keep_tmp = TRUE` leaves the files in `tempdir()` until they are manually
removed or the R session ends. Repeated or parallel searches can retain large
amounts of data, potentially exhausting temporary storage or a user quota.
Use `keep_tmp = FALSE` for routine work.

Common failures can be diagnosed as follows:

- **Executable not found:** inspect `Sys.which(required_tools)` and correct
  `PATH`.
- **Database does not exist:** pass the common database prefix, **without** an
  index extension such as `.nhr`, `.nin`, or `.nsq`.
- **FASTA write error:** check available space, permissions, and
  quotas for `tempdir()`.
- **Nonzero BLAST exit status:** repeat one query with `verbose = TRUE` and
  `keep_tmp = TRUE`, then inspect the printed command and retained files.
- **Incomplete output:** first resolve any preceding FASTA or BLAST error;
  parsing messages are often downstream symptoms of a failed command.

On systems with a small `/tmp`, select a larger temporary filesystem before
starting R:

```{bash, eval = FALSE}
TMPDIR=/path/with/space Rscript analysis.R
```

# Prebuilt NCBI databases

`blast_db_get()` downloads and caches archives from the
[NCBI BLAST database directory](https://ftp.ncbi.nlm.nih.gov/blast/db/).
Because these archives can be large and require network access, this example is
not executed while building the vignette:

```{r prebuilt-database, eval = run_local, eval = FALSE}
archive <- blast_db_get("16S_ribosomal_RNA.tar.gz")
database_dir <- "/path/to/16S_rRNA_DB"
dir.create(database_dir, recursive = TRUE, showWarnings = FALSE)
untar(archive, exdir = database_dir)

ncbi_db <- blast(file.path(database_dir, "16S_ribosomal_RNA"))
hits <- predict(ncbi_db, queries[1])
```

Large databases may be divided into numbered archives. Download and extract
every part into the same directory. See the
[BLAST command-line manual](https://www.ncbi.nlm.nih.gov/books/NBK279690/)
for database layouts and search options.

# Remote searches

A BLAST object can refer to an NCBI database without downloading it:

```{r remote-search, eval = run_local, eval = FALSE}
remote_db <- blast("nt", remote = TRUE, type = "blastn")
remote_hits <- predict(remote_db, queries[1])
```

Remote searches use a shared NCBI service, may be queued, and are usually much
slower than local searches. They also require network access. Prefer a local
database for large or repeated analyses.

```{r cleanup, include = FALSE, eval = run_local}
unlink(work_dir, recursive = TRUE)
```

# Session information

```{r session-info}
sessionInfo()
```
